%---------------------------------------------
% This document is for pdflatex
%---------------------------------------------
\documentclass[11pt]{article}

\usepackage{amsmath,amsfonts,amssymb,graphicx,natbib,setspace,authblk}
\usepackage[running]{lineno}
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{rotating}

\usepackage[compact]{titlesec}

\usepackage{graphicx}

\usepackage{times}

\usepackage{setspace}

\usepackage{enumitem}
\setlist{topsep=.125em,itemsep=-0.15em,leftmargin=0.75cm}

\usepackage[textsize=footnotesize,textwidth=2.1cm]{todonotes}
\newcommand{\smalltodo}[2][]
    {\todo[caption={#2}, #1]
    {\begin{spacing}{0.5}#2\end{spacing}}}
% There is nothing \smalltodo{here is a comment} here.
% \todo[inline]{Here is a comment}

\renewcommand{\floatpagefraction}{0.98}
\renewcommand{\topfraction}{0.99}
\renewcommand{\textfraction}{0.05}

\clubpenalty = 10000
\widowpenalty = 10000

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{equation} \begin{aligned}}
\newcommand{\ea}{\end{aligned} \end{equation}}

\def\X{\mathbf{X}}
\def\A{\mathbf{A}}
\def\B{\mathbf{B}}
\def\C{\mathbf{C}}
\def\D{\mathbf{D}}
\def\G{\mathbf{G}}
\def\H{\mathbf{H}}
\def\N{\mathbf{N}}
\def\M{\mathbf{M}}
\def\W{\mathbf{\Omega}}
\def\P{\mathbf{P}}
\def\V{\mathbf{V}}

\title{Detecting and predicting the relationship between climate and population dynamics}

\author[1]{Andrew T. Tredennick\thanks{Corresponding author: atredenn@gmail.com}}
\author[1]{Peter B. Adler}
\author[2]{Giles Hooker}
\author[3]{Stephen P. Ellner}
\author[1]{Brittany J. Teller}
\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan Utah}
\affil[2]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
\affil[3]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
\renewcommand\Authands{, and }

\date{Last compile: \today}

\renewcommand{\baselinestretch}{1.2}
\linespread{1.5}

\begin{document}

\maketitle

\vspace{1in} 

\large
{\emph{Running head}: Predicting climate effects}
\normalsize 

\newpage

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
\linenumbers

\section*{Abstract}
Populations fluctuate in response to internal and external forces.
Quantifying the influence of external forces on population dynamics is a classic focus of population ecology, and it has taken on new relevance as we face the challenge of predicting species' responses to climate change.
However, detecting climate--population growth relationships is difficult and, once detected, the relationship may not improve out of sample prediction --- the gold standard of forecasting.
Here, we review empirical studies to assess how often including climate, or climate-related, covariates improves out of sample prediction of population abundances and dynamics.
We find that...
Drawing on examples from our literature review and our own work (failures?), we outline ways to determine in advance whether one can predict out of sample.
We then use a simulation approach to show how to identify the most efficient ways to improve predictive skill and, in turn, allocate research effort. 
 

\subsection*{Keywords:} Prediction, climate, forecast, population, validation  

\newpage

\section*{Introduction}
Why populations fluctuate has fascinated generations of ecologists.
Do populations fluctuate because of strong internal regulation and density-dependence?
Or, do external forces, like weather, determine population growth and decline from year-to-year?
Finding answers to these questions is intellectually satisfying, and doing so has taken on new relevance as we confront the challenge of predicting the impacts of climate change.
Whether our motivation is fundamental or applied, discriminating between internal and external population regulation requires quantifying the effects of external covariates on population growth, either directly or through demographic rates.

Unfortunately, detecting such climate--population growth relationships is not easy.
One issue is the high deminsionality of the problem: there are generally a handfull of plausible environmental or climate covariates that can potentially affect a number of fitness components, and each covariate must be aggregated over some time period to be tractable in a statistical model.
Another issue is that few population time series are long enough and replicated enough to meet the demands of estimating more than a couple covariate effects (Teller et al. 2016).
In other words, it is nontrivial to decide which covariates to use and difficult to detect significant effects of those covariates given data limitations.

Even when detecting a significant climate--population growth relationship is possible, doing so does not always lead to significantly better predictions.





\end{document}
